Abstractions and exemplars: the measure noun phrase alternation in German

### 2018-04-21 Latex/sections/3corpus.tex
Second half of proofreading.

For the present study, I used the German \textit{Corpus from the Web} (COW) in its 2014 version DECOW14A (\citealp{SchaeferBildhauer2012full}, and \citealp{Schaefer2015b}, as well as \citealp{BiemannEa2013}, and \citealp{SchaeferBildhauer2013}, for overviews of web corpora in general and the methodology of their construction), which contains almost 21 billion tokens.%
\footnote{The COW corpora (Dutch, English, French, German, Spanish, Swedish) are made available for free at \url{https://www.webcorpora.org}.
At the time of this writing, a newer 2016 version DECOW16A has already been released.}
I chose this corpus for two main reasons.
First, the external validity of any study is increased through a higher heterogeneity of the sample \citep[30]{MaxwellDelaney2004}, and the DECOW14A corpus has clearly a much more heterogeneous composition compared to the only other very large corpus of German, the DeReKo of the Institute for the German Language \citep{KupietzEa2010}, which contains almost exclusively newspaper texts.%
\footnote{It was shown in \cite{W16-2601} that, for example, the range of topics is much smaller in DeReKo compared to DECOW14A.}
Second, it was already mentioned that normative grammars often adopt clear positions regarding the grammaticality of either the \NACa\ or the \PGCa.
Thus, newspaper text or any other text that conforms strongly to normative grammars might not represent the alternation phenomenon fully and without bias because authors and proofreaders who must adhere to normative guidelines might favour one alternative or the other explicitly.
Web corpora, on the other hand, contain a significant amount of non-standard language from forums and similar sources.
For such reasons, COW corpora have been used in a number of peer-reviewed publications, for example \cite{VanGoethemHiligsmann2014}, \cite{VanGoethemHuening2015}, \cite{MuellerS2014}, \cite{Schaefer2016c}, \cite{SchaeferSayatz2014}, \cite{SchaeferSayatz2016}, and \cite{Zimmer2015}. 
Therefore, DECOW14A is a valid choice for this study.

@@

Among the factors potentially influencing the alternation (see Section~\ref{sec:germanmeasurenps}) are lemma-specific effects.
Therefore, it was desirable to obtain a sample in which most of the highly frequent actually-occurring combinations of kind nouns and measure nouns were represented.
I applied a two-stage process in order to obtain a list of co-occurring measure nouns and kind nouns.

In the first step, I exported a list of all nouns from the DECOW14A01 sub-corpus (one billion tokens large).
Then, the one hundred most frequent mass nouns were extracted manually from this list.
Mass nouns were defined as concrete nouns which genuinely denote a substance, combine with uninflected mass quantifiers such as \textit{viel} `much' and \textit{wenig} `little' (\textit{viel Bier} `much beer'), and form only sortal and unit plurals (such as the plural \textit{Biere} `types of beer' or `glasses of beer').
Abstract nouns which partially behave like mass nouns (like \textit{Spaß} `fun’ or \textit{Gefahr} `danger’) were excluded because they are not measured in the same way as concrete mass nouns.

@@

To capture the influence of style mentioned in Section~\ref{sec:analyses}, two proxy variables were used.
At the document level, the DECOW14A corpus has an annotation for \textit{Badness}.
As described in \cite{SchaeferEa2013}, \textit{Badness} measures how well the distribution of highly frequent short words in the document matches a language model of standard German.
The paper also shows that the Badness score corresponds robustly with human raters' intuitions about text coherence and text quality.
Documents with higher Badness usually contain more incoherent language and shorter sentences.
If the \PGCa\ actually favours more elaborate stylistic levels, a high \textit{Badness} should be correlated with fewer occurrences.
Documents in DECOW14 have also been annotated with a variable called \textit{Genitives}.
The higher the values of this variable, the lower the proportion of genitives among all case-bearing forms is.
A high number of genitives is also indicative of a more formal, elaborate style close to the written standard.
However, since the \PGCa\ contains a genitive itself, the regressor variable \textit{Genitive} and the document-level variable \textit{Genitives} are not fully independent.
Since instances of the \PGCa\ make up for only a minute fraction of all genitives, I still use \textit{Genitives} as a regressor with the appropriate caveats.

@@

Finally, two variables were added as nuisance variables in the context of the present study.
First, it was reported in the literature that MNPs in the dative and with a masculine or neuter kind noun favour the \PGCa\ more than the corresponding nominative and accusative MNPs \citep{Hentschel1993,Zimmer2015}.
As an example, \textit{mit einem Stück frischen Brots} `with a piece of fresh bread' (\PGCa) would be preferred more strongly against \textit{mit einem Stück frischem Brot} (\NACa).
As with all the examples, native speakers of German will most likely notice that differences are subtle.
The case of the measure noun was manually annotated (variable \textit{Measurecase}).
Second, due to differences in case syncretisms, it is likely that feminine kind nouns have slightly different preferences than masculine and neuter ones, and hence the appropriate variable \textit{Kindgender} was annotated.

@@

A multilevel logistic regression model was fit which models the influence of the regressors specified in Table~\ref{tab:variables} on the probability that the \PGCa\ is chosen over the \NACa.
All regressors from Table~\ref{tab:variables} were included, and the measure lemma and the kind noun lemma were specified as varying-intercept random effects.
The sample size was \textit{n=}5,063 with 1,134 cases of \PGCa\ and 3,929 cases of \NACa.
The results of the estimation are shown in Figure~\ref{fig:fixeffs} and in Table~\ref{tab:bigtable}.
The intercept comprises \textit{Cardinal=Yes}, \textit{Measurecase=Nom}, \textit{Kindgender=Masc}, \textit{Measureclass=Physical}, and 0 for all numeric z-transformed regressors.
It was estimated at -3.548.

@@

Using signed logarithmised Fisher p-values as a measure of collexeme strength (\textit{Measurecollo} and \textit{Kindcollo}) instead of the quotient for the attraction strength (\textit{Measureattraction} and \textit{Kindattraction}) (see Section~\ref{sec:variablesandannotation}) was not successful.
The $p_{\text{PB}}$ value for \textit{Kindcollo} was 0.191 and the one for \textit{Measurecollo} was 0.443.
The coefficients of determination dropped to $R_m^2=0.376$ and $R^2_c=0.480$.

@@

The results generally confirm the hypotheses from Section~\ref{sec:analyses}.
First, the exemplar effect (the influence of the non-alternating \PGCd\ and \NACb) was shown (see the effect plots in Figure~\ref{fig:eff:attraction}).%
\footnote{Effect plots were created using the \textit{effects} package \citep{Fox2003}.}
The effect is as expected:
if a lemma appears relatively more often in the \PGCd\ (compared to its frequency in the \NACb), the \PGCa\ tends to be chosen over the \NACa\ with this specific lemma.
The effect for measure nouns is stronger, and it was estimated with higher precision.

>>>>

For the present study, I used the 21 billion token German \textit{Corpus from the Web} (COW) in its 2014 version DECOW14A (\citealp{SchaeferBildhauer2012full}, and \citealp{Schaefer2015b}, as well as \citealp{BiemannEa2013}, and \citealp{SchaeferBildhauer2013}, for overviews of web corpora in general and the methodology of their construction).%
\footnote{The COW corpora (Dutch, English, French, German, Spanish, Swedish) are made available for free at \url{https://www.webcorpora.org}.
At the time of this writing, a newer 2016 version DECOW16A has already been released.}
This corpus was chosen first because the external validity of any study is increased through a more heterogeneous sample \citep[30]{MaxwellDelaney2004}.
DECOW14A clearly has a much more heterogeneous composition compared with the only other very large corpus of German, the DeReKo of the Institute for the German Language \citep{KupietzEa2010}.
DeReKo contains almost exclusively newspaper texts.%
\footnote{It was shown in \cite{W16-2601} that, for example, the range of topics is much smaller in DeReKo than in DECOW14A.}
The second reason for choosing this particular corpus is that normative grammars often adopt clear positions regarding the grammaticality of either the \NACa\ or the \PGCa.
Thus, newspaper text or any other texts which conform strongly to normative grammars may not represent the alternation phenomenon fully and without bias because authors and proofreaders who must adhere to normative guidelines might explicitly favour one alternative over another.
Web corpora, on the other hand, contain a significant amount of non-standard language from forums and similar sources.
For such reasons, COW corpora have been used in a number of peer-reviewed publications \citep{VanGoethemHiligsmann2014,VanGoethemHuening2015,MuellerS2014,Schaefer2016c,SchaeferSayatz2014,SchaeferSayatz2016,Zimmer2015}.

@@

Among the factors potentially influencing the alternation (see Section~\ref{sec:germanmeasurenps}) are lemma-specific effects.
Therefore, it was necessary to obtain a sample in which most of the highly frequent actually-occurring combinations of kind nouns and measure nouns were represented.
Hence, I applied a two-stage process in order to obtain a list of co-occurring measure nouns and kind nouns.

In the first step, I exported a list of all nouns from the DECOW14A01 sub-corpus (one billion tokens large).
Then, the one hundred most frequent mass nouns were extracted manually from this list.
Mass nouns were defined as concrete nouns which genuinely denote a substance, combine with uninflected mass quantifiers such as \textit{viel} `much' and \textit{wenig} `little' (\textit{viel Bier} `much beer'), and form only sortal and unit plurals (such as the plural \textit{Biere} `types of beer' or `glasses of beer').
Abstract nouns which partially behave as mass nouns (for example, \textit{Spaß} `fun’ or \textit{Gefahr} `danger’) were excluded because they are not measured in the same way as concrete mass nouns.

@@

To capture the influence of style mentioned in Section~\ref{sec:analyses}, two proxy variables were used.
At the document level, the DECOW14A corpus has an annotation for \textit{Badness}.
As described in \cite{SchaeferEa2013}, \textit{Badness} measures how well the distribution of highly frequent short words in the document matches a language model of standard German.
The paper also shows that the Badness score corresponds robustly with human raters' intuitions about text coherence and text quality.
Documents with higher Badness usually contain more incoherent language and shorter sentences.
If the \PGCa\ actually favours more elaborate stylistic levels, a high \textit{Badness} should be correlated with fewer occurrences.
Documents in DECOW14 have also been annotated with a variable called \textit{Genitives}.
For each document it is the case that the higher the values of this variable, the lower the proportion of genitives among all case-bearing forms.
A high number of genitives is also indicative of a more formal, elaborate style close to the written standard.
However, since the \PGCa\ contains a genitive itself, the regressor variable \textit{Genitive} and the document-level variable \textit{Genitives} are not fully independent.
Since instances of the \PGCa\ make up for only a minute fraction of all genitives, I still use \textit{Genitives} as a regressor with the appropriate caveats.

@@

Finally, two variables have been added as nuisance variables in the context of the present study.
First, the case of the measure noun was manually annotated (variable \textit{Measurecase}).
This was carried out because it was reported in the literature that MNPs in the dative containing a masculine or neuter kind noun favour the \PGCa\ more than the corresponding nominative and accusative MNPs \citep{Hentschel1993,Zimmer2015}.
As an example, \textit{mit einem Stück frischen Brots} `with a piece of fresh bread' (\PGCa) would be preferred more strongly against \textit{mit einem Stück frischem Brot} (\NACa).
As with all the examples, native speakers of German will most likely notice that differences are subtle.
Second, due to differences in case syncretisms, it is likely that feminine kind nouns have slightly different preferences than masculine and neuter ones, and hence the appropriate variable \textit{Kindgender} was annotated.

@@

A multilevel logistic regression model was fit which models the influence of the regressors specified in Table~\ref{tab:variables} on the probability that the \PGCa\ is chosen over the \NACa.
All regressors from Table~\ref{tab:variables} were included, and the measure lemma and the kind noun lemma were specified as varying-intercept random effects.
The sample size was \textit{n=}5,063 with 1,134 cases of \PGCa\ and 3,929 cases of \NACa.
The results of the estimation are shown in Figure~\ref{fig:fixeffs} and in Table~\ref{tab:bigtable}.
The intercept estimated at -3.548 comprises \textit{Cardinal=Yes}, \textit{Measurecase=Nom}, \textit{Kindgender=Masc}, \textit{Measureclass=Physical}, and 0 for all numeric z-transformed regressors.

@@

The use of signed logarithmised Fisher p-values as a measure of collexeme strength (\textit{Measurecollo} and \textit{Kindcollo}) instead of the quotient for the attraction strength (\textit{Measureattraction} and \textit{Kindattraction}) (see Section~\ref{sec:variablesandannotation}) was not successful.
The $p_{\text{PB}}$ value for \textit{Kindcollo} was 0.191 and the one for \textit{Measurecollo} was 0.443.
The coefficients of determination dropped to $R_m^2=0.376$ and $R^2_c=0.480$.

@@

The results generally confirm the hypotheses from Section~\ref{sec:analyses}.
First, the exemplar effect (the influence of the non-alternating \PGCd\ and \NACb) was validated (see the effect plots in Figure~\ref{fig:eff:attraction}).%
\footnote{Effect plots were created using the \textit{effects} package \citep{Fox2003}.}
The effect is as expected:
if a lemma appears relatively more often in the \PGCd\ (compared to its frequency in the \NACb), the \PGCa\ tends to be chosen over the \NACa\ with this specific lemma.
The effect for measure nouns is even stronger and could be estimated with higher precision.

### 2018-04-21 Latex/sections/4experiments.tex
Second half of proofreading.

The decision to use attested material and perform a global validation of the corpus-based model was not arbitrary.
Alternatively, one could have created stimuli where the features favouring the alternatives were permuted and thus tested directly.
The answer is related to what was said in Section~\ref{sec:cogocl} about the rich multi-factorial data sets used in corpus studies and the comparatively restricted ones in experiments.
Looking at Table~\ref{tab:variables}, we see one binary factor (\textit{Cardinal}), two three-level factors (\textit{Measurecase} and \textit{Kindgender}) and one five-level factor (\textit{Measureclass}).
Since the dependent variable is binary (\textit{Construction}), permuting these factors alone leads to 180 different possible permutations.
Additionally, the numeric variables \textit{Kindattraction} and \textit{Measureattraction} would also have to be tested at several values.
It is obvious that this is impossible in a controlled experiment, given that a participant can be exposed to no more than roughly one hundred sentences, most of which would be fillers rather than target sentences.
Thus, while the present approach only allows for a global test of the corpus-derived model, this appears to me to be the only feasible way.%
\footnote{For the same reasons, the experimental data are also much too sparse to perform post-hoc analyses with respect to the single regressors in a sound way.}
It is, in the sense of Section~\ref{sec:cogocl}, an optimal synthesis of data-rich multifactorial corpus studies and experimental validation, and it is one that has been used at least since the seminal \citet{BresnanEa2007} paper -- although rarely, as \citet[3--4]{DivjakEa2016a} point out.

@@

The final pairs of stimuli were the sentence containing the attested and preferred alternant (according to the corpus GLMM) on the one hand and a modified version containing the dispreferred alternant on the other hand.
They were presented next to each other, and a 20 second time limit for each choice was set.%
\footnote{No participant ever exceeded the time limit.}
The order of sentences was randomised for each participant, and the position of the alternants on the screen (left\slash right) was randomised per participant and sentence.
As fillers, 23 pairs of sentences exemplifying similar but unrelated alternation phenomena from German morpho-syntax were used.
Thus, participants saw 39 pairs of sentences and 78 sentences in total.
They were instructed to select from each pair of sentences the one that seemed more natural to them in the sense that they would use it rather than the other one.
The experiment was conducted using \textit{PsychoPy} \citep{Peirce2007}.

>>>>

The decision to use attested material and perform a global validation of the corpus-based model was not arbitrary.
Alternatively, one could have created stimuli where the features favouring the alternatives were permuted and thus tested directly.
The answer is related to the statement in Section~\ref{sec:cogocl} about the rich multi-factorial data sets used in corpus studies and the comparatively restricted ones in experiments.
Looking at Table~\ref{tab:variables}, we see one binary factor (\textit{Cardinal}), two three-level factors (\textit{Measurecase} and \textit{Kindgender}) and one five-level factor (\textit{Measureclass}).
Since the dependent variable is binary (\textit{Construction}), permuting these factors alone leads to 180 different possible permutations.
Additionally, the numeric variables \textit{Kindattraction} and \textit{Measureattraction} would also have to be tested at several values.
It is obvious that this is impossible in a controlled experiment, given that a participant can be exposed to no more than roughly one hundred sentences, most of which would be fillers rather than target sentences.
Thus, while the present approach only allows for a global test of the corpus-derived model, this appears to me to be the only feasible way.%
\footnote{For the same reasons, the experimental data are also much too sparse to perform post-hoc analyses with respect to the single regressors in a sound way.}
It is, in the sense of Section~\ref{sec:cogocl}, an optimal synthesis of data-rich multifactorial corpus studies and experimental validation, and it is one that has been used at least since the seminal \citet{BresnanEa2007} paper.
As \citet[3--4]{DivjakEa2016a} pointed out, however, the numbers of published studies including validation is still inadequate.

@@

The final pairs of stimuli were the sentences containing the attested and preferred alternant (according to the corpus GLMM) and a modified version containing the dispreferred alternant.
They were presented next to each other, and a 20 second time limit for each choice was set.%
\footnote{No participant ever exceeded the time limit.}
The order of sentences was randomised for each participant, and the position of the alternants on the screen (left\slash right) was randomised per participant and sentence.
As fillers, 23 pairs of sentences exemplifying similar but unrelated alternation phenomena from German morpho-syntax were used.
Thus, participants saw 39 pairs of sentences and 78 sentences in total.
They were instructed to select from each pair of sentences the one that seemed more natural to them in the sense that they would use it rather than the other one.
The experiment was conducted using \textit{PsychoPy} \citep{Peirce2007}.
